ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data

### 2021-06-29 manuscript/supplement.tex
update manuscript to 2nd revision

The eventual goal in variant calling is to infer the true frequency of a possible alternative, and thus variant, nucleotide in the sample you sequenced.
For such an alternative nucleotide $x\in\{A,C,G,T\}$ at a particular site in a genome, we therefor define\footnote{We omit an explicit notation for a site, but every $\theta$ is always specific to one genome position.}:
\begin{equation}
 \theta_{b}(x), \theta_{s}(x)
\end{equation}
as the (true, but usually unknown) frequency of alternative nucleotide $x$ in the bulk and single cell sample, respectively.
These frequencies translate into probabilities of sampling a DNA fragment containing the respective nucleotide at a site:
either directly, as with the sequencing of DNA from a bulk of cells $b$;
or indirectly, as with the sequencing of whole genome amplified DNA of a single cell $s$.
It is our eventual goal to work back from bulk and single cell data to derive likelihoods for possible values of $\theta_{b}(x)$ and $\theta_{s}(x)$, respectively, allowing us to place likelihoods on possible underlying genotypes and events like allele dropouts in single cells.\\

For a single diploid cell $s$, the possible alternative nucleotide frequencies are
\begin{equation}
 \theta_s(x) =
 \begin{cases}
  0       \\
  \frac12 \\
  1
 \end{cases}
\end{equation}
reflecting the cases that:
(i) $x$ is not present at the position---if no other allele is present, this constitutes a homozygous reference site;
(ii) $x$ reflects an alternative nucleotide that is present in only one of the two genome copies of the single cell---this constitutes a heterozgyous site;
(iii) $x$ is an alternative nucleotide present in both genome copies of the single cell---this constitutes a homozygous alternative site.

@@

@@

For the current implementation, we use the empirical values for $\alpha(l), \beta(l), w(l), \alpha_1(l), \alpha_2(l)$ as deduced by \cite{lodato_somatic_2015} (Table~\ref{tab:Lodato}).
But as mentioned above, these could be learned from a dataset at hand whenever high-confidence sets of heterozygous and homozygous sites are available (e.g.~by calling variants on a background bulk and cross-referencing with dbSNP).\\

@@

Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell and the bulk sequencing sample, given the respective data.
All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequency likelihood density estimates $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
We then define mutually exclusive sub-spaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$, providing a set of events $\mathfrak{E}$.
For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:event-space}, lower panel).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
For the above example ($\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$), this means that we can resolve a contradiction between the likelihood estimate at $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and the likelihood density estimate across $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
Based on our assumption, we decide to trust the bulk sample over the single cell sample, meaning that the bulk is a mixture of a maximum of two sub-populations, and can only contain heterozygous and homozygous alternative cells, rendering a homozygous reference single cell impossible and thus the result of an allele dropout.

@@

\textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
For each, 16 cells and a bulk sample were whole exome sequenced.
For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
Depending on the single cell, we used two different ground truths for two separate evaluations:
\begin{enumerate}
    \item For the tumor single cells, we constructed a ground truth from normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
    Thus, similar caveats as for the germline ground truth in the granulocyte dataset apply.
    To mitigate this to a certain degree, we added the 373 clonal validated somatic variants, as these are expected to be present in all tumor single cells.
    \item For the normal single cells, we constructed a ground truth from the tumor bulk sample (as the normal bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    To reduce the presence of false-positive somatic variants in this ground truth, we removed all 467 validated somatic variants (clonal and subclonal) from the ground truth.
\end{enumerate}
Further, we determined the of the validated somatic variants for all software run on this dataset.

@@

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=42ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

@@

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf}
 \caption{
 Cell-specific $F_1$ scores for different filtering threshold levels for the whole genome dataset.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:F_1-scores_Dong2017}
\end{figure}

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_F1_plot.pdf}
 \caption{
 Cell-specific $F_1$ scores for different filtering threshold levels for the whole exome dataset.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:F_1-scores_Laehnemann2017}
\end{figure}

ProSolo's improvements in accuracy can be further formalised using the $F_1$ score, the harmonic mean of precision $p$ and recall $r$, with $\beta = 1$ for the general $F$ score defined as \citep{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}:\\

\begin{equation} \label{eq:F_1_score}
  F_{\beta} = (\beta^2 + 1) \times \frac{p r}{\beta^2 p + r}
\end{equation}

The per-cell $F_1$ score clearly demonstrates ProSolo's performance advantage (Figures~\ref{fig:F_1-scores_Dong2017} and \ref{fig:F_1-scores_Laehnemann2017}), even when restricting our comparison to ProSolo's (and SCIPhI's) default mode for a fairer comparison on the whole exome dataset with a germline ground truth (see caveat discussed above).

@@

For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
On the normal cells, both ProSolo and SCIPhI achieve a precision above .99.
Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a drastically precision.
This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
Where ProSolo models each cell and each genomic site separately, SCIPhI's model integrates information across all sites in all cells at once.
While this can clearly help recover recall, this can also lead to false positive somatic variant calls (Figure~\ref{fig:somatic-recall}) and has clear implications for model complexity and software runtime.
As all other tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
However, it should be noted that SCIPhI took from 1 week (with iterations reduced below software defaults) on the single nucleus whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, this will further increase SCIPhI's wall time.
In addition, with more cells added, the space of possible tree topologies that SCIPhI explores grows exponentially, further increasing runtimes.

@@

Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
For ProSolo, this amounts to the default mode without imputation (which also ensures a fair comparison against the germline-based ground truth) and controlling for a false discovery rate of $0.2$.
For MonoVar, this means no consensus filtering and no filtering via the {\ttfamily --t} thresholding parameter.\\

>>>>

The eventual goal in variant calling is to infer the true frequency of a possible alternative, and thus variant, nucleotide in the sample you sequenced.
For such an alternative nucleotide $x\in\{A,C,G,T\}$ at a particular site in a genome, we therefor define\footnote{We omit an explicit notation for a site, but every $\theta$ is always specific to one genome position.}:
\begin{equation}
 \theta_{b}(x), \theta_{s}(x)
\end{equation}
as the true (but usually unknown) frequency of alternative nucleotide $x$ in the bulk and single cell sample, respectively.
These frequencies translate into probabilities of sampling a DNA fragment containing the respective nucleotide at a site:
either directly, as with the sequencing of DNA from a bulk of cells $b$;
or indirectly, as with the sequencing of whole genome amplified DNA of a single cell $s$.
It is our eventual goal to work back from bulk and single cell data to derive likelihoods for all possible values of $\theta_{b}(x)$ (denoted as $\tilde{\theta_b}(x)$)\footnote{We use $\theta$ notation for the true underlying alternate allele frequency at a site, and $\tilde{\theta}$ notation when referring to the likelihood density estimates derived for all possible underlying alternate allele frequencies. The latter notation is thus mostly used for the definition of single cell events in section \ref{sec:event-space-def}.} and $\theta_{s}(x)$ (denoted as $\tilde{\theta_s}(x)$), respectively, allowing us to place likelihoods on possible underlying genotypes and events like allele dropouts in single cells.\\

For a single diploid cell $s$, the true alternative nucleotide frequency can be
\begin{equation}
 \theta_s(x) =
 \begin{cases}
  0       \\
  \frac12 \\
  1
 \end{cases}
\end{equation}
reflecting the cases that:
(i) $x$ is not present at the position---if no other allele is present, this constitutes a homozygous reference site;
(ii) $x$ reflects an alternative nucleotide that is present in only one of the two genome copies of the single cell---this constitutes a heterozgyous site;
(iii) $x$ is an alternative nucleotide present in both genome copies of the single cell---this constitutes a homozygous alternative site.

@@

\begin{table}[tbp]
 \caption{
 Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
 Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.
 }
 \label{tab:Lodato-full}
 \renewcommand{\arraystretch}{1.15}

 \begin{center}
  \begin{tabular}{lrr}
   parameter  & slope        & intersect   \\
   \toprule
   $\alpha$   & -0.000027183 & 0.068567471 \\
   $\beta$    & 0.007454388  & 2.367486659 \\
   \midrule
   $w$        & 0.000548761  & 0.540396786 \\
   $\alpha_1$ & 0.057378844  & 0.669733191 \\
   $\alpha_2$ & 0.003233912  & 0.399261625 \\
   \bottomrule
  \end{tabular}
 \end{center}
\end{table}

@@

For the current implementation, we use the empirical values for $\alpha(l), \beta(l), w(l), \alpha_1(l), \alpha_2(l)$ as deduced by \cite{lodato_somatic_2015} (Table~\ref{tab:Lodato-full}).
But as mentioned above, these could be learned from a dataset at hand whenever high-confidence sets of heterozygous and homozygous sites are available (e.g.~by calling variants on a background bulk and cross-referencing with dbSNP).\\

@@

\subsubsection{Defining the Event Space}\label{sec:event-space-def}

Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell ($\tilde{\theta_s}$) and the bulk sequencing sample ($\tilde{\theta_b}$), given the respective data.
All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequency likelihood density estimates $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
We then define mutually exclusive sub-spaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$, providing a set of events $\mathfrak{E}$.
For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:event-space}, lower panel).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
For the above example ($\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$), this means that we can resolve a contradiction between the likelihood estimate at $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and the likelihood density estimate across $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
Based on our assumption, we decide to trust the bulk sample over the single cell sample, meaning that the bulk is a mixture of a maximum of two sub-populations, and can only contain heterozygous and homozygous alternative cells, rendering a homozygous reference single cell impossible and thus the result of an allele dropout.

To evaluate whether this assumption of only two sub-populations in the bulk is safe to make, we can examine all the possible ways in which this assumption may be violated.
This requires that we have a sequence of two mutational events affecting the same genomic site, each generating a new sub-population.
To know which types of mutational event sequences to consider, we note that the same point mutation has been observed to occur independently in different single cells from the same individual \citep{kuipers_single-cell_2017}, possibly due to selection for an advantageous effect of that mutation.
However, the net effect of such parallel evolution still restricts the cell bulk to two sub-populations with distinct genotypes\footnote{As we are looking at different single cells separately, this also means that we do not have to worry about this kind of violation of the infinite sites assumption. Such parallel evolution only becomes problematic for phylogenetic reconstruction approaches that assume infinite sites, which would assume that each mutation only happens once in the tree.}.
At the same time, we are not aware of cases where different cells from the same population acquire different point mutations at the exact same locus, and expect this to be highly unlikely.
Instead, a third sub-population will usually be generated by a loss of heterozygosity event on a larger scale, covering a mutated site \citep{kuipers_single-cell_2017,satas_scarlet_2020}.

Thus, if we start with a homozygous reference bulk, a first event would introduce an alternative allele mutation and a second event in that sub-population would lead to the loss of either the reference allele or the alternative allele.
Losing the recently generated alternative allele leads to an allele frequency in that sub-population that is similar to the homozygous reference genotype (it only has coverage of the reference allele), and this sub-population is thus indistinguishable from the original homozygous reference population.
ProSolo will correctly identify any single cells with the loss event as homozygous reference (Figure~\ref{fig:event-space}).

However, losing the reference allele leads to a sub-population that only has coverage of the alternative allele and is thus similar to a homozygous alternative genotype and clearly distinguishable from the other two genotypes.
While this type of event was not observed in a systematic survey of multiple mutational hits per site \citep{kuipers_single-cell_2017}, it has coincidentally been observed in a recent study that explicitly modeled copy number alterations in single cells for their phylogenetic reconstruction of cell lineages (mutation LINGO2:2 in \cite{satas_scarlet_2020}).
Thus, we expect this type of event to be rare, but possible.
In this case, as the alternative allele was only introduced in a sub-population and the loss of the reference allele is only present in an even smaller sub-population of that, the bulk frequency of this alternative allele will be well below $0.5$.
Here, the ProSolo model assumes the homozygous alternative genotype of this sub-population impossible and instead calls cells from that sub-population as heterozygous (Figure~\ref{fig:event-space}).
However, this only concerns the genotype, while the presence of the mutation will be identified correctly.
As MonoVar is the only other variant calling software that can even genotype beyond calling the presence or absence of a mutation, this could only be considered a drawback compared to that tool.

As germline homozygous alternative sites are very rare compared to germline homozygous reference sites (due to the reference genome being biased towards more common alleles), we expect the symmetric case\footnote{A mutation towards the reference with a subsequent complete loss of the originally homozygous alternative allele.} to be much less likely to be observed.

Finally, starting from a heterozygous germline genotype in the bulk will be more likely than starting from a homozygous alternative genotype, but a lot less likely than starting from a homozygous reference genotype.
From this starting point, three sub-populations regarding the same alternative allele can only be generated by two events in opposite directions in distinct cells from that population:
one point mutation or loss event towards the reference and one point mutation or loss event towards the alternative allele, with the loss events much more likely to affect any site than the point mutations.
Whether the resulting bulk alternative allele frequency then ends up above or below $0.5$, depends on which event happens earlier and/or grows faster within the population.
As a result, the homozygous genotype (reference or alternative) with the lower frequency will erroneously be considered a heterozygous single cell genotype, but ProSolo will correctly call the more frequent (and thus probably more relevant) homozygous genotype.
However, we are not aware of any such events being reported in the literature and expect them to be vary rare.

We conclude that the assumption is reasonable to make and we only expect it to lead to the misclassification of homozygous alternative single cell sites as heterozygous when they have been generated by the introduction of an alternative allele via a point mutation, followed by an allelic loss of the reference allele in some daughter cells.
We also discuss possible respective cases in the context of the tumor single cells from the data of \citet{wang_clonal_2014} (Supplement Section~\ref{sec:somatic-recall}).

It should be possible to mitigate such cases with further extensions of the model to account for copy number changes.
As in \citet{satas_scarlet_2020}, this would require separately obtaining copy number profiles that are then provided as input to ProSolo, but we do not yet see a standardized way of obtaining such profiles from single cell DNA-sequencing data.
So far, such copy number profiles seem to have usually been obtained with custom scripts tailored to specific datasets (for example \cite{baslan_genome-wide_2012,satas_scarlet_2020,kuipers_single-cell_2020}), but when more generalized approaches for obtaining such profiles become available, they could then be used to adjust the exact model setup for each site based on the local copy number profile.
Newer versions of varlociraptor, whose variant calling functionality ProSolo uses in the background, already allow for a chromosome-specific specification of copy numbers\footnote{See the "ploidy" definition in this documentation section: \url{https://varlociraptor.github.io/docs/calling/\#configuring-the-joint-prior-distribution}}.
Thus, it should be possible to extend this to site-specific copy numbers in the future and to extend the ProSolo model to arbitrary ploidies.

\subsubsection{Posterior Probabilities of Events}\label{sec:event-posteriors}

@@

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_validated_somatic_clonal_zygosity_misclassifications.pdf}
 \caption{
   The number of clonal variants with a respective alternative allele frequency, stratified by the genotype (hom for homozygous alternative and het for heterozygous) provided by \cite{wang_clonal_2014} for their somatic ground truth list.
   We consider the peak around $0.4$ to represent genuine heterozygous variants and the peak around $0.7$ to represent genuine homozygous alternative variants, with a tumor sample purity in the duplex sequencing experiment of somewhere around those $0.7$.
 }
\label{fig:wang-clonals-zygosity}
\end{figure}

\textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
For each, 16 cells and a bulk sample were whole exome sequenced (Figure~\ref{fig:datasets}C).
For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
As they provide a duplex sequencing based alternative allele frequency, we examined the alternative allele frequency distribution of the non-synonymous clonal variants, stratified by the specified zygosity (Figure~\ref{fig:wang-clonals-zygosity}).
This clearly shows two alternative allele frequency peaks, one at around $0.4$ and the other at roughly double the alternative allele frequency around $0.7$.
These two peaks should correspond to heterozygous ($0.4$) and homozygous alternative variants ($0.7$), respectively.
As these are supposed to be clonal variants in the tumor, the remaining fraction of $0.3$ beyond the homozygous alternative variants peak (from $0.7$ to $1$) should be contamination with normal cells (without these somatic variants).
This indicates a duplex sample tumor purity of around $0.7$.
Unexpectedly, we also see the $0.7$ peak in variants that \cite{wang_clonal_2014} classified as heterozygous and the $0.4$ peak in variants that \cite{wang_clonal_2014} classified as homozygous alternative (Figure~\ref{fig:wang-clonals-zygosity}).
It is unclear from \cite{wang_clonal_2014} and its supplementary material how the original variant classifications were generated, we assume that this is based on the genotypes obtained for the single cells.
But as the single cells were genotyped with standard variant calling software (GATK UnifiedGenotyper), they very likely suffer from considerable allelic dropout and we thus reclassified the clonal variants' genotypes based on the duplex sequencing.
If the alternative allele frequency determined by the ultra-deep duplex sequencing was below $0.6$, we classified the variant as heterozygous, and as homozygous alternative if this frequency was above (or equal to) $0.6$ (Figure~\ref{fig:wang-clonals-zygosity}).
Then, depending on the single cell, we used two different ground truths for two separate evaluations:
\begin{enumerate}
    \item For the tumor single cells, we constructed a ground truth from the normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
    Thus, similar caveats as for the germline ground truth in the granulocyte dataset apply.
    To mitigate this to a certain degree, we added the 373 clonal validated somatic variants, as these are expected to be present in all tumor single cells.
    \item For the normal single cells, we constructed a ground truth from the tumor bulk sample (as the normal bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    To reduce the presence of false-positive somatic variants in this ground truth, we removed all 467 validated somatic variants (clonal and subclonal) from the ground truth.
\end{enumerate}
Further, we determined the of the validated somatic variants for all software run on this dataset.

@@

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=42ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

@@

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_per-cell_prec-rec.pdf}
 \caption{
 Cell-specific precision-recall plots across different filtering threshold levels for the whole genome dataset.\newline \footnotesize
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:per-cell_prec-rec_Dong2017}
\end{figure}

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_per-cell_prec-rec.pdf}
 \caption{
 Cell-specific precision-recall plots across different filtering threshold levels for the whole exome dataset.\newline \footnotesize
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline
  Used thresholds are not comparable between tools.
  They were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:per-cell_prec-rec_Laehnemann2017}
\end{figure}

@@

For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
On the normal cells, both ProSolo and SCIPhI achieve a precision above .99.
Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a reduced precision.
This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
Where ProSolo models each cell and each genomic site separately, SCIPhI's model integrates information across all sites in all cells at once.
While this can clearly help recover recall, this can also lead to false positive somatic variant calls (Figure~\ref{fig:somatic-recall}) and has clear implications for model complexity and software runtime.
As all other tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
However, it should be noted that SCIPhI took from 1 week (with iterations reduced below software defaults) on the single nucleus whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, this will further increase SCIPhI's wall time.
In addition, with more cells added, the space of possible tree topologies that SCIPhI explores grows exponentially, further increasing runtimes.

@@

Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C), with the $F_1$ score defined as the harmonic mean of precision $p$ and recall $r$ (with $\beta = 1$ for the general $F$ score, see \cite{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}):\\

\begin{equation} \label{eq:F_1_score}
  F_{\beta} = (\beta^2 + 1) \times \frac{p r}{\beta^2 p + r}
\end{equation}.

For ProSolo, this amounts to the default mode without imputation (which also ensures a fair comparison against the germline-based ground truth) and controlling for a false discovery rate of $0.2$.
For MonoVar, this means no consensus filtering and no filtering via the {\ttfamily --t} thresholding parameter.\\
